A Large-Scale Study of Polymorphism in R

### 2019-04-05 paper/main.tex
method fixes and edits

In this section, we detail our methodology for collecting data.  Our aim is
to observe arguments and return values of function calls, and from these
generalize possible type signatures for the called functions.  We base our
infrastructure on an open source tool called \genthat whose purpose is to
generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
by synthesizing unit tests from recorded function argument and return
values, comparing the test against existing ones to avoid generating tests
which do not increase code coverage.  To suit our purposes, we change the
existing tool in two main ways: we record \emph{shapes} rather than values,
and ignore the code coverage optimization phase.  Both changes are
beneficial for scalability, allowing us to trace far more calls.

@@

\begin{figure}[!hb]
\begin{tabular}{ll}\begin{minipage}{5cm}
{\small\begin{lstlisting}[style=R]
> 1 + 2L
[1] 3
> x <- c(1,2,3,4)
> attr(x,"dim") <- c(2,2)
> x + c(1,2)
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}
\end{minipage} &
\begin{minipage}{8cm}\small
\begin{tabular}{rl}
\tt `+`: &\tt \sD \sI \to \sD \\
\tt c:& \sD \sD \sD \sD \to \D\\
\tt `<-`: &\tt \sY \D \to \D\\
\tt c: & \tt \sD \sD \to \D\\
\tt `attr<-`:&\tt  \sY \C \D \to \\
\tt c: &\tt \sD \sD \to \D\\
\tt `+`: &\tt \attr\D{dim=\D} \D \to \attr\D{dim=\D}
\end{tabular}
\end{minipage}
\end{tabular}
\caption{Example script and recorded signature}\label{example}\end{figure}

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}.  The shapes we would expect to record are as follows:
In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
\k{character}. For vectorized types we record their length and for all types
we record attributes with some of their values. Looking at the signatures
observed for addition, it is clear that the function is polymorphic as it
starts with the addition of two scalar numbers of different types, and then
adds a matrix to a vector returning a matrix.

@@

\begin{enumerate}
\item {\bf Exec generation:} All runnable code in the package is extracted
  from its tests, examples and vignettes. The code snippets are combined
  into a single file.
\item {\bf Installation:} All packages that are required for execution of
  the current package are downloaded and installed.
\item {\bf Instrumentation:} As code is loaded into the R, every function
  definition is instrumented with an \code{on-exit} hook which is invoked
  when the function returns either normally or through an exception.
\item {\bf Recording:} When a hook is called, arguments and return value are
  inspected. We record \k{typeof}, \k{class} and \k{attributes} recursively.
  For \code{list} values, an extra bit of analysis is performed to record
  the element type.
\item {\bf Writing:} Unique signatures are recorded to file with information
  about which package triggered the recording.
\end{enumerate}

Our recording mechanism does not remember the order in which arguments were
passed, nor does it record which arguments were not passed (and for which
default values were used). If an argument was not passed, and no default
value was specified, we record it as \emph{missing}. In R, trying to use
such a missing value results in an error. For practical reasons, we do not
record the contents of environments. These can be used as hash tables and
may be big and are quite likely different from one another.

In R all arguments are passed as promises, and unused arguments will be
unevaluated.  Our tracer does not collect information on arguments that were
not evaluated as they may not be valid expressions.  Consider the {\tt
  magrittr} package, which implements the pipe function that performs a kind
of currying.

@@

The call to the pipe desugars to \code{`\%>\%`(lhs=4, rhs=mul(3))} where the
\code{rhs} is not a valid expression. It is treated as code that will
manipulated to add one argument (the \code{lhs}).

To be as specific as possible, our analysis collects shape information about
arguments and return values. The details of the information we collect follows.

\begin{itemize}
\item If the analysis encounters a primitive vector, say
  \code{typeof(x)=="double"}, it records the length of {\tt x} to determine
  if it can be treated as a scalar.
\item For a list, \code{typeof(x)=="list"}, shape information on all
  elements is used to describe the list's shape.  To avoid undue slowdowns,
  we only collect the content's \code{typeof}, and ascribe \ANY (any) if the
  elements are not of the same kind.k
\item If the analysis encounters a scalar \code{NA}, we ascribe a unique
  NULL type.  In R, \code{NA} inhabits all types, but for a scalar NA,
  \code{typeof(NA)=="logical"}. Scalar NAs are used as uninitialized values.
\item For a matrix (i.e. the value has matrix class and has a dims and
  optionally dimnames attribute), we ascribe the shape \attr{\tt T}{mat}
  where {\tt T} is the primitive element type.
\item For a data frame (i.e., class {\tt data.frame} with appropriate
  attributes), we ascribe the shape \attr{\l}{df}.
\end{itemize}

@@

For a meaningful notion of polymorphism, it is necessary to settle on a set
of basic types for R. There are several possible choices which lead to
increasingly more expressive type systems. The simplest notion of types
aligns with R's builtin types. Richers notions incorporate attributes and
the various object systems built on top of those. For concreteness we will
pick out three different sets of types.

>>>>

In this section, we detail our methodology for collecting data.  Our aim is
to observe arguments and return values of function calls, and from these
generalize possible type signatures for the called functions.  We base our
infrastructure on an open source tool called \genthat whose purpose is to
generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
by synthesizing unit tests from recorded function argument and return
values, comparing candidate tests against existing ones to avoid generating tests
which do not increase code coverage.  To suit our purposes, we change the
existing tool in two main ways: we record \emph{shapes} rather than values,
and ignore the code coverage optimization phase.  Both changes are
beneficial for scalability, allowing us to trace far more calls.

@@

\begin{figure}[!hb]
\begin{tabular}{ll}\begin{minipage}{5cm}
{\small\begin{lstlisting}[style=R]
> 1 + 2L
[1] 3
> x <- c(1,2,3,4)
> attr(x,"dim") <- c(2,2)
> x + c(1,2)
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}
\end{minipage} &
\begin{minipage}{8cm}\small
\begin{tabular}{rl}
\tt `+`: &\tt \sD \sI \to \sD \\
\tt c:& \sD \sD \sD \sD \to \D\\
\tt `<-`: &\tt \sY \D \to \D\\
\tt c: & \tt \sD \sD \to \D\\
\tt `attr<-`:&\tt  \sY \C \D \to \\
\tt c: &\tt \sD \sD \to \D\\
\tt `+`: &\tt \attr\D{dim} \D \to \attr\D{dim}
\end{tabular}
\end{minipage}
\end{tabular}
\caption{Example script and recorded signature}\label{example}\end{figure}

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}, and the shapes we expect to record are given to the right of the figure.
We abbreviate the types \k{double}, \k{integer}, \k{symbol} and \k{character} to \sD, \sI, \sY, and \sC for simplicity.
\D, \I, \Y, and \C denote vectors, whereas \sD, \sI, \sY, and \sC denote scalars.
We record the lengths of all vectorized-typed values, and for all types we record their classes and attribute names. 
Looking at the signatures observed for addition, it is clear that the function is polymorphic: it starts with the addition of two scalar numbers of different types, and later adds a matrix to a vector returning a matrix.

@@

\begin{enumerate}
\item {\bf Exec generation:} All runnable code in the package is extracted
  from its tests, examples and vignettes. The code snippets are combined
  into a single file.
\item {\bf Installation:} All packages that are required for execution of
  the current package are downloaded and installed.
\item {\bf Instrumentation:} As code is loaded into the R, every function
  definition is instrumented with an \code{on-exit} hook which is invoked
  when the function returns either normally or through an exception.
\item {\bf Recording:} When a hook is called, arguments and return value are
  inspected. We record \k{typeof}, \k{class} and \k{attributes} recursively.
  For \code{list} values, an extra bit of analysis is performed to record
  the element types.
\item {\bf Writing:} Unique signatures are recorded to file with information
  about which package triggered the recording.
\end{enumerate}

In R, all arguments are passed as promises, and unused arguments will be unevaluated. 
Our recording mechanism captures the order in which arguments are passed, can figure out which arguments were {\it not} provided (recording them as {\it missing} if the missing argument has no default value), and does {\it not} force unforced promises, recording unforced promises as {\it unevaled}.
In R, trying to use missing values results in an error and thus an exception, triggering our recording phase: arguments which cause errors in this way are recorded as {\it error}.
For practical reasons, we do not record the contents of environments:
These can be used as hash tables, may be large, and are quite likely different from one another.

Unevaluated arguments in R are sometimes {\it carefully} used by their functions to implement some form of metaprogramming.
Consider the {\tt magrittr} package, which implements a pipe function \%$>$\% that implements a flavor of function currying:

@@

The call to the pipe desugars to \code{`\%>\%`(lhs=4, rhs=mul(3))} where the
\code{rhs} is not a valid expression. It is treated as code that will
manipulated to add one argument (the \code{lhs}), resulting in the function
actually performing a call of the form \code{mul(4, 3)}.

As we alluded to at the beginning of this section, our analysis collects shape information about
arguments and return values in order to be as specific as possible. The details of the information we collect follows.
Note that the first step of any type collection in our analysis begins with \code{typeof}.

\begin{itemize}
\item If the analysis encounters a primitive vector, say
  \code{typeof(x)=="double"}, it records the length of {\tt x} to determine
  if it can be treated as a scalar.
\item For a list, \code{typeof(x)=="list"}, shape information on all
  elements is used to describe the list's shape.  To avoid undue slowdowns,
  we only collect the content's \code{typeof}, and ascribe \ANY (any) if the
  elements are not of the same kind.
\item If the analysis encounters a scalar \code{NA}, we ascribe a unique
  NULL type.  In R, \code{NA} inhabits all types, but for a scalar NA,
  \code{typeof(NA)=="logical"}. Scalar NAs are used as uninitialized values.
\item For a matrix, \code{typeof(m)=="some\_primitive\_type"}, and\code{m} will have matrix class and also a dims and
  optionally dimnames attribute. Here, we ascribe the shape \attr{\tt T}{mat}
  where {\tt T} is the primitive element type.
\item For a data frame, \code{typeof(df)=="list"}, and \code{df} will have class {\tt data.frame}, and {\tt names} and {\tt row.names} attributes.
 To reflect this, we ascribe the shape \attr{\l}{df}.
\end{itemize}

@@

For a meaningful notion of polymorphism, it is necessary to settle on a set
of basic types for R. There are several possible choices which lead to
increasingly more expressive type systems. The simplest notion of types
aligns with R's builtin types. Richer notions incorporate attributes and
the various object systems built on top of those. For concreteness we will
pick out three different sets of types.
